Hilbert space approximate Bayesian Gaussian processes using Stan

### 2020-04-10 Paper/manuscript_v2/manuscript_07_01.tex
Abstract and Sec. 1

\usepackage{listings}
\usepackage{slashbox}
\usepackage{caption}
\captionsetup{font=footnotesize}

@@

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions, however they entail a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov chain Monte Carlo. In this paper, we focus on a novel approach for low-rank approximate Bayesian Gaussian processes, based on a basis functions approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance and practical implementation of the method in relation to key factors such as the number of basis functions, domain of the prediction space, and smoothness of the latent function. We provide intuitive visualizations and recommendations for choosing the values of these factors, which make it easier for users to improve approximation accuracy and computational performance. We also propose a diagnostics for checking that the number of basis functions and the domain of the prediction space are adequate given the data. The proposed approach is simple and exhibits an attractive computational complexity due to its linear structure, and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\end{abstract}

@@

Sparse GPs are based on low-rank approximations of the covariance matrix. The low-rank approximation with $m \ll n$ {\it inducing points} implies reduced memory requirements of $O(nm)$ and corresponding computational complexity of $O(nm^2)$.
A unifying view on sparse GPs based on approximate generative methods
is provided by \citet{quinonero2005unifying}, while a general review
is provided by \citet{rasmussen2006gaussian}. \citet{Burt+Rasmussen+vanderWilk:2019} show that for regression with normally distributed covariates in $D$ dimensions and using the squared exponential covariance function, $M=O(\log^DN)$ is sufficient for an accurate approximation.
% More recent developments in the context of sparse GPs include a structured kernel interpolation method \citep{wilson2015kernel}. More recently, \citep{bui2017unifying} proposed to use inducing points-based sparse approximation methods to perform GP approximations at inference time rather than at modeling time.

An alternative class of low-rank approximations is based on forming a basis function approximation with $m \ll n$ basis functions. The basis functions are usually presented explicitly, but can also be used to form a low rank covariance matrix approximation. Common basis function approximations rest on the spectral analysis and series expansions of GPs \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}.
Sparse spectrum GPs are based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving}. Recently, \citet{hensman2017variational} presented a variational Fourier feature approximation for GPs that was derived for the Mat{\'e}rn class of kernels. Another related method for approximating kernels relies on random Fourier features \citep{rahimi2008random,rahimi2009weighted}.
Certain spline smoothing basis functions are equivalent to GPs with certain covariance functions \citep{wahba1990spline,Furrer+Nychka:2007}.
A recent related work based on a spectral representation of GPs as an infinite series expansion with the Karhunen-Loève representation \citep[see, e.g.][]{grenander1981abstract} is presented by \citet{JSSv090i10}.
% , but not all spline models are GPs (e.g., some spline models are not marginalization consistent).

@@

The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance function including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost of learning the covariance function parameters is $O(mn+m)$ in every step of the optimizer or sampler. This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentiation methods used in modern probabilistic programming frameworks, such as Stan \citep{carpenter2017stan} and others. This is because the memory requirements of automatic differentiation scale with the size of the autodiff expression tree which in direct implementations is simpler for basis function than covariance matrix based approach. \todo{Better?}. The basis function approach also provides an easy way to apply a non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

@@

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks. In this paper, the interactions among the key factors of the method such as the number of basis functions, domain of the prediction space, and properties of the true functional relationship between covariates and response variable, are investigated and analyzed in detail in relation to the computational performance and accuracy of the method. Practical recommendations are given for the values of the key factors based on intuitive graphical summaries that encode the recognized relationships. Our recommendations will help users to choose valid and optimized values for these factors, improving computational performance without sacrificing modeling accuracy. We propose also diagnostics to indicate whether the chosen values for the number of basis functions and the domain of the prediction space are adequate to model the data well.

@@

We have implemented the approach in the probabilistic programming language Stan \citep{carpenter2017stan} as well as subsequently in the \textit{brms} package \citep{burkner2017brms} of the R software \citep{R2019R}. Several illustrative examples of the performance and applicability of the method are shown using both simulated and real datasets. All examples are accompanied by the corresponding Stan %and R 
code.

Although there are several GP specific software packages available to date, for example, GPML \citep{rasmussen2010gpml},  GPstuff \citep{vanhatalo2013gpstuff}, GPy \citep{gpy2014}, and GPflow \citep{GPflow2017}), each provide efficient implementations only for a restricted range of GP-based models. In this paper, we do not focus on the fastest possible inference for some specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks.

The remainder of the paper is structured as follows. In Section \ref{ch4_gp}, we introduce GPs, covariance functions and spectral density functions. In Section \ref{ch5_sec_method}, the reduced rank approximation to GPs proposed by \citet{solin2018hilbert} is described. In Section \ref{ch5_sec_accuracy}, the accuracy of these approximations under several conditions using analytical and numerical methods is analyzed. Several case studies in which we fit
exact and approximate GPs to real and simulated data are provided in Section \ref{ch5_sec_cases}. A brief conclusion of the work is made in Section \ref{ch5_sec_conclusion}. Appendix \ref{ch5_app_approx_covfun} includes a brief presentation of the mathematical details behind the Hilbert space approximation of a stationary covariance function, and Appendix \ref{ch5_sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function. 
Online supplemental material with more case studies illustrating the performance and applicability of the method can be found at \url{https://github.com/gabriuma/basis_functions_approach_to_GP} in the subfolder \url{Paper/online_supplemental_material}.

>>>>

\usepackage{listings}
%\usepackage{slashbox}
\usepackage{caption}
\captionsetup{font=footnotesize}

\usepackage[colorlinks=true]{hyperref}

@@

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions. However they entail a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov chain Monte Carlo. In this paper, we focus on a novel approach for low-rank approximate Bayesian Gaussian processes, based on a basis function approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance and practical implementation of the method in relation to key factors such as the number of basis functions, domain of the prediction space, and smoothness of the latent function. We provide intuitive visualizations and recommendations for choosing the values of these factors, which make it easier for users to improve approximation accuracy and computational performance. We also propose diagnostics for checking that the number of basis functions and the domain of the prediction space are adequate given the data. The proposed approach is simple and exhibits an attractive computational complexity due to its linear structure, and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\end{abstract}

@@

Sparse GPs are based on low-rank approximations of the covariance matrix. The low-rank approximation with $m \ll n$ {\it inducing points} implies reduced memory requirements of $O(nm)$ and corresponding computational complexity of $O(nm^2)$. A unifying view on sparse GPs based on approximate generative methods is provided by \citet{quinonero2005unifying}, while a general review is provided by \citet{rasmussen2006gaussian}. \citet{Burt+Rasmussen+vanderWilk:2019} show that for regression with normally distributed covariates in $D$ dimensions and using the squared exponential covariance function, $M=O(\log^DN)$ is sufficient for an accurate approximation.
% More recent developments in the context of sparse GPs include a structured kernel interpolation method \citep{wilson2015kernel}. More recently, \citep{bui2017unifying} proposed to use inducing points-based sparse approximation methods to perform GP approximations at inference time rather than at modeling time.
An alternative class of low-rank approximations is based on forming a basis function approximation with $m \ll n$ basis functions. The basis functions are usually presented explicitly, but can also be used to form a low-rank covariance matrix approximation. Common basis function approximations rest on the spectral analysis and series expansions of GPs \citep{loeve1977probability,trees1968detection,adler1981geometry,cramer2013stationary}. Sparse spectrum GPs are based on a sparse approximation to the frequency domain representation of a GP \citep{lazaro2010sparse,quia2010sparse,gal2015improving}. Recently, \citet{hensman2017variational} presented a variational Fourier feature approximation for GPs that was derived for the Mat{\'e}rn class of kernels. Another related method for approximating kernels relies on random Fourier features  \citep{rahimi2008random,rahimi2009weighted}. Certain spline smoothing basis functions are equivalent to GPs with certain covariance functions \citep{wahba1990spline,Furrer+Nychka:2007}. Recent related work based on a spectral representation of GPs as an infinite series expansion with the Karhunen-Loève representation \citep[see, e.g.,][]{grenander1981abstract} is presented by \citet{JSSv090i10}.
% , but not all spline models are GPs (e.g., some spline models are not marginalization consistent).

@@

The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance function including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost of learning the covariance function parameters is $O(mn+m)$ in every step of the optimizer or sampler. This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentiation methods used in modern probabilistic programming frameworks, such as Stan \citep{carpenter2017stan} and others. This is because the memory requirements of automatic differentiation scale with the size of the autodiff expression tree which in direct implementations is simpler for basis function than covariance matrix based approach. The basis function approach also provides an easy way to apply a non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

@@

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks. In this paper, the interactions among the key factors of the method such as the number of basis functions, domain of the prediction space, and properties of the true functional relationship between covariates and response variable, are investigated and analyzed in detail in relation to the computational performance and accuracy of the method. Practical recommendations are given for the values of the key factors based on intuitive graphical summaries that encode the recognized relationships. Our recommendations will help users to choose valid and optimized values for these factors, improving computational performance without sacrificing modeling accuracy. We also propose  diagnostics to indicate whether the chosen values for the number of basis functions and the domain of the prediction space are adequate to model the data well.

@@

We have implemented the approach in the probabilistic programming language Stan \citep{carpenter2017stan} as well as subsequently in the \textit{brms} package \citep{burkner2017brms} of the R software \citep{R2019R}. Several illustrative examples of the performance and applicability of the method are shown using both simulated and real datasets. All examples are accompanied by the corresponding Stan code. Although there are several GP specific software packages available to date, for example, GPML \citep{rasmussen2010gpml},  GPstuff \citep{vanhatalo2013gpstuff}, GPy \citep{gpy2014}, and GPflow \citep{GPflow2017}, each provide efficient implementations only for a restricted range of GP-based models. In this paper, we do not focus on the fastest possible inference for some specific GP models, but instead are interested in how GPs can be easily used as modular components in probabilistic programming frameworks.

The remainder of the paper is structured as follows. In Section~\ref{ch4_gp}, we introduce GPs, covariance functions and spectral density functions. In Section~\ref{ch5_sec_method}, the reduced-rank approximation to GPs proposed by \citet{solin2018hilbert} is described. In Section~\ref{ch5_sec_accuracy}, the accuracy of these approximations under several conditions using analytical and numerical methods is analyzed. Several case studies in which we fit exact and approximate GPs to real and simulated data are provided in Section~\ref{ch5_sec_cases}. A brief conclusion of the work is made in Section~\ref{ch5_sec_conclusion}. Appendix~\ref{ch5_app_approx_covfun} includes a brief presentation of the mathematical details behind the Hilbert space approximation of a stationary covariance function, and Appendix \ref{ch5_sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function. Online supplemental material with more case studies illustrating the performance and applicability of the method can be found at \url{https://github.com/gabriuma/basis_functions_approach_to_GP} in the subfolder \url{Paper/online_supplemental_material}.
